Some examples in extended logic counting classes

### 2016-12-27 LICS/counting.tex
Cambios en las citas.

\marcelo{Ahora solo consideramos la idea de capturar sobre estructuras ordenadas}

@@

\marcelo{Hacer una pasada a la demostracion del siguiente teorema considerando que solo usamos estructuras ordenadas, y teniendo en cuenta que las funciones estan definidas sobre relational signatures.}
\martin{done}
\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}

@@

Saluja et. al \cite{DBLP:journals/jcss/SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\so$. For a formula $\varphi(x,X)$ over $\R$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\ostr[\R]$. A function $f:\Sigma^*\to\nat$ is in $\#\LL$ if there exists an $\LL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$. Let $\sqso(\LL)$ be a fragment of $\qso$ defined by the following grammar:
\[
\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha,
\]
where $\varphi$ is an $\LL$-formula. It can be seen that $\sqso(\LL)$ captures $\#\LL$ over ordered structures and that $\sqso(\LL) \subseteq \eqso(\LL)$.

>>>>

@@

\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}

@@

Saluja et. al \cite{SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\so$. For a formula $\varphi(x,X)$ over $\R$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\ostr[\R]$. A function $f:\Sigma^*\to\nat$ is in $\#\LL$ if there exists an $\LL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$. Let $\sqso(\LL)$ be a fragment of $\qso$ defined by the following grammar:
\[
\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha,
\]
where $\varphi$ is an $\LL$-formula. It can be seen that $\sqso(\LL)$ captures $\#\LL$ over ordered structures and that $\sqso(\LL) \subseteq \eqso(\LL)$.

### 2016-12-27 LICS/saluja.tex

Inspired by the connection between $\shp$ and $\sfo$, the authors in~\cite{DBLP:journals/jcss/SalujaST95} studied subclasses of  $\sfo$ syntactically restricting the alternation quantifiers in boolean formulas defining what we call the $\sfo$-hierarchy.
Specifically, they define the classes $\E{i}$ and $\U{i}$ where functions are defined by $\fo$-formulas in $\loge{i}$ and $\logu{i}$, respectively, for every $i \geq 0$. 
Interestingly, in~\cite{DBLP:journals/jcss/SalujaST95} they show that this function classes defined a finite hierarchy of the form:
\[
\E{0} \; \subsetneq \; \E{1} \; \subsetneq \; \U{1} \; \subsetneq \; \E{2} \; \subsetneq \; \U{2} \; = \; \#\fo 
\]
In this section, we study a natural extension of these classes in light of the $\eqso$ logic. Formally, we define the classes $\eqso(\loge{i})$ and $\eqso(\logu{i})$ where the boolean logic is restricted to $\loge{i}$ and $\logu{i}$, respectively. 
We denote each class by $\QE{i}$ and $\QU{i}$ for short.

In the next result, we show that $\eqso$ also defined a finite hierarchy similar than in~\cite{DBLP:journals/jcss/SalujaST95} that we called the $\eqso$-hierarchy.
\begin{proposition}
\[
\QE{0} \; \subsetneq \; \QE{1} \; \subsetneq \; \QU{1} \; \subsetneq \; \QE{2} \; \subsetneq \; \QU{2} \; = \; \eqso{\fo} 
\]
\end{proposition}
A natural question until this point is what is the connection between $\E{i}$ and $\U{i}$ and their corresponding classes $\QE{i}$ and $\QU{i}$. Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ or $\U{i}$ is equivalent to $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ which is in $\QE{i}$ or $\QU{i}$, respectively. 
Then it left to know whether the containment between the sharp and quantitative classes is strict or not. 
For this, we study first whether each formula $\alpha$ in $\eqso(\LL)$ can be transformed in a \emph{prefix normal form} where all sums quantifiers are restricted to the beginning of the formula. 
Formally, we say that a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\Sigma$-prenex normal form} (or just prenex normal form) if $\alpha$ is of the form:
\[
\alpha := \sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})
\]
where $\bar{X}$ and $\bar{x}$ is a sequence of $0$ or more second and first order variables, respectively.
In particular, note that a boolean formula is always in $\Sigma$-prenex normal form.
Then we say that a formula $\alpha$ is in \emph{sum normal form} if $\alpha$ is of the form $c + \Sigma_{i=1}^n \alpha_i$ where $c$ is a non-negative constant and each $\alpha_i$ is in $\Sigma$-prenex normal form.
Next we show that each formula in $\eqso(\LL)$ can be converted in \emph{sum normal form} but not always in prenex normal form.
\begin{theorem}
	Every formula in $\eqso(\LL)$ can be rewritten in \emph{sum normal form}. Furthermore, there exists a formula $\alpha \in \loge{1}$ such that there does not exist a formula in $\Sigma$-prenex normal form in $\QE{1}$ equivalent to $\alpha$.
\end{theorem}

@@

The previous results gives the connection the connection between the hierarchy in \cite{DBLP:journals/jcss/SalujaST95} and the $\eqso$-hierarchy. 
Indeed, if $\LL$ contains $\logu{1}$ we will have that $\sh{\LL}$ is equivalent to $\eqso(\LL)$ since each formula in $\eqso(\LL)$ can be converted in $\Sigma$-prefix normal form and, therefore, it is equivalent to a formula in $\sh{\LL}$. 
Unfortunately, this is not the case for $\loge{0}$ and $\loge{1}$ as the following result shows.
\begin{proposition}
	The classes $\E{0}$ and $\E{1}$ are strictly contained in $\QE{0}$ and $\QE{1}$, respectively. Moreover, the classes $\U{1}$, $\E{2}$, and $\U{2}$ are equivalent with $\QU{1}$, $\QE{2}$, and $\QU{2}$, respectively.
\end{proposition}
The previous result shows that the classes $\QE{i}$ and $\QU{i}$ are more robust than the classes $\E{i}$ and $\U{i}$: they are closed under binary and sum quantifiers but the other not necessarily.

@@

%We see that many of the results in Saluja et. al. \cite{DBLP:journals/jcss/SalujaST95} for $\#\LL$ still apply in $\eqso(\LL)$ for a given fragment $\LL$:
%
%\begin{theorem} \label{eqso-sigma-zero-in-fp}
%	For every $\eqso(\loge{0})$ formula $\alpha$ over a signature $\R$, the function $f$ over $\R$ defined as $f(\enc(\A)) = \sem{\alpha}(\A)$ is in $\fp$.
%\end{theorem}
%
%\begin{theorem} \label{eqso-sigma-one-in-eqso-pi-one}
%	For every $\eqso(\loge{1})$ formula $\alpha$ over a signature $\R$ there exists a $\eqso(\logu{1})$ formula $\beta$ over $\R$ such that $\sem{\alpha}(\A) = \sem{\beta}(\A)$ for every $\A\in\ostr[\R]$.
%\end{theorem}
%
%
%The {\em decision problem} associated to a function $f$ is defined by the language $L_f = \{\A \in \str \mid f(\A) > 0\}$.
%
%\begin{theorem} \label{decisionptime}
%	The decision problem associated to a function in $\eqso(\logex{1})$ is in \textsc{P}.
%\end{theorem}

>>>>

Inspired by the connection between $\shp$ and $\sfo$, the authors in~\cite{SalujaST95} studied subclasses of  $\sfo$ syntactically restricting the alternation quantifiers in boolean formulas defining what we call the $\sfo$-hierarchy.
Specifically, they define the classes $\E{i}$ and $\U{i}$ where functions are defined by $\fo$-formulas in $\loge{i}$ and $\logu{i}$, respectively, for every $i \geq 0$. 
Interestingly, in~\cite{SalujaST95} they show that this function classes defined a finite hierarchy of the form:
\[
\E{0} \; \subsetneq \; \E{1} \; \subsetneq \; \U{1} \; \subsetneq \; \E{2} \; \subsetneq \; \U{2} \; = \; \#\fo 
\]
In this section, we study a natural extension of these classes in light of the $\eqso$ logic. Formally, we define the classes $\eqso(\loge{i})$ and $\eqso(\logu{i})$ where the boolean logic is restricted to $\loge{i}$ and $\logu{i}$, respectively. 
We denote each class by $\QE{i}$ and $\QU{i}$ for short.

In the next result, we show that $\eqso$ also defined a finite hierarchy similar than in~\cite{SalujaST95} that we called the $\eqso$-hierarchy.
\begin{proposition}
\[
\QE{0} \; \subsetneq \; \QE{1} \; \subsetneq \; \QU{1} \; \subsetneq \; \QE{2} \; \subsetneq \; \QU{2} \; = \; \eqso{\fo} 
\]
\end{proposition}
A natural question until this point is what is the connection between $\E{i}$ and $\U{i}$ and their corresponding classes $\QE{i}$ and $\QU{i}$. Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ or $\U{i}$ is equivalent to $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ which is in $\QE{i}$ or $\QU{i}$, respectively. 
Then it left to know whether the containment between the sharp and quantitative classes is strict or not. 
For this, we study first whether each formula $\alpha$ in $\eqso(\LL)$ can be transformed in a \emph{prefix normal form} where all sums quantifiers are restricted to the beginning of the formula. 
Formally, we say that a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\Sigma$-prenex normal form} (or just prenex normal form) if $\alpha$ is of the form:
\[
\alpha := \sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})
\]
where $\bar{X}$ and $\bar{x}$ is a sequence of $0$ or more second and first order variables, respectively.
In particular, note that a boolean formula is always in $\Sigma$-prenex normal form.
Then we say that a formula $\alpha$ is in \emph{sum normal form} if $\alpha$ is of the form $c + \Sigma_{i=1}^n \alpha_i$ where $c$ is a non-negative constant and each $\alpha_i$ is in $\Sigma$-prenex normal form.
Next we show that each formula in $\eqso(\LL)$ can be converted in \emph{sum normal form} but not always in prenex normal form.
\begin{theorem}
	Every formula in $\eqso(\LL)$ can be rewritten in \emph{sum normal form}. Furthermore, there exists a formula $\alpha \in \loge{1}$ such that there does not exist a formula in $\Sigma$-prenex normal form in $\QE{1}$ equivalent to $\alpha$.
\end{theorem}

@@

The previous results gives the connection the connection between the hierarchy in \cite{SalujaST95} and the $\eqso$-hierarchy. 
Indeed, if $\LL$ contains $\logu{1}$ we will have that $\sh{\LL}$ is equivalent to $\eqso(\LL)$ since each formula in $\eqso(\LL)$ can be converted in $\Sigma$-prefix normal form and, therefore, it is equivalent to a formula in $\sh{\LL}$. 
Unfortunately, this is not the case for $\loge{0}$ and $\loge{1}$ as the following result shows.
\begin{proposition}
	The classes $\E{0}$ and $\E{1}$ are strictly contained in $\QE{0}$ and $\QE{1}$, respectively. Moreover, the classes $\U{1}$, $\E{2}$, and $\U{2}$ are equivalent with $\QU{1}$, $\QE{2}$, and $\QU{2}$, respectively.
\end{proposition}
The previous result shows that the classes $\QE{i}$ and $\QU{i}$ are more robust than the classes $\E{i}$ and $\U{i}$: they are closed under binary and sum quantifiers but the other not necessarily.

@@

%We see that many of the results in Saluja et. al. \cite{SalujaST95} for $\#\LL$ still apply in $\eqso(\LL)$ for a given fragment $\LL$:
%
%\begin{theorem} \label{eqso-sigma-zero-in-fp}
%	For every $\eqso(\loge{0})$ formula $\alpha$ over a signature $\R$, the function $f$ over $\R$ defined as $f(\enc(\A)) = \sem{\alpha}(\A)$ is in $\fp$.
%\end{theorem}
%
%\begin{theorem} \label{eqso-sigma-one-in-eqso-pi-one}
%	For every $\eqso(\loge{1})$ formula $\alpha$ over a signature $\R$ there exists a $\eqso(\logu{1})$ formula $\beta$ over $\R$ such that $\sem{\alpha}(\A) = \sem{\beta}(\A)$ for every $\A\in\ostr[\R]$.
%\end{theorem}
%
%
%The {\em decision problem} associated to a function $f$ is defined by the language $L_f = \{\A \in \str \mid f(\A) > 0\}$.
%
%\begin{theorem} \label{decisionptime}
%	The decision problem associated to a function in $\eqso(\logex{1})$ is in \textsc{P}.
%\end{theorem}
